---
title: "Gov 50 Final Project"
author: "Your Name Here"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Project thoughts

---- milestone 1: 

I am interested in exploring data related to identity, specifically the Hispanic/Latinx identity in the United States.This data set from the Pew Research Center analyzes how individuals with Hispanic/Latinx ancestry do or do not identify with the terms. 

This is of particular interest to me as a dual citizen with Brazil (where my grandfather is from). I myself tend to identify as white -- especially if the only two options are 'white, not hispanic/latino' or 'hispanic/latino', and if there is no overlap between the two. I would generally tend to identify more with Brazil than with any Latin identity, especially as I am racially white. I think it is relatively common among Latin/Hispanic people to identify most strongly with orgin country... thus I will also take origin country into account in this study. 

I would like to examine how individuals WITH HISPANIC/LATIN heritage identify racially. I will only look at participants with some sort of hispanic/latin ancestry. I'm thinking of having three main variables: race (hispanic or latino vs. not hispanic/latino); hispanic/latin ancestry (either yes or no); and grouping of origin country. 

Those with hispanic/latin heritage who identify as hispanic/latin will be considered "self identifying". Those with only ancestry of, but no racial self-identification, will be what I hope to explain through examining this dataset. 


---- milestone 2: 


My research question is... 

Individuals with hispanic and latinx heritage vary in their self identification in being hispanic and/or latinx. Does this vary across generations -- in their separation from their ancestors -- and is there variation across countries that might also influence it? 

dependent variable: self-identifying as hispanic or latinx
independent variables: degree of ancestry (were you born there, your parents, your grandparents) + country of ancestry 

These are measured based on questions on a survey, all in a yes/no format, with the exception of the country of ancestry (which is a fill in the bubble with options of all hispanic/latinx countries). 

Hypothesis: Those who were born in a latinx or hispanic country AND those who have 2 parents that are hispanic/latinx will identify more strongly. Those with only one parent will do so less, along with those with 2+ levels of separation. In terms of origin countries, I believe that that will only have a strong impact after 2 levels of separation. At which point, only people from countries that have a strong immigrant prescence in the US will continue to associate strongly with hispanic/latinx heritage. 

https://www.pewresearch.org/hispanic/dataset/2016-survey-of-self-identified-non-hispanics-with-hispanic-ancestry/




```{r}
library(haven)
library(sjlabelled)
library(tidyverse)
library(infer)

pew <- read_sav("Data/Pew/NSL2015 Omnibus_FOR RELEASE.sav") |>
  sjlabelled::as_label()

```

```{r}
nationality <- pew |>
  select(q1, q3_combo) |>
  mutate(q1 = ifelse(q1 == "No",0,1)) |>
  count(q3_combo) 
number_each_country
knitr::kable(nationality,
             col.names = c("Nationality", "Count"))
```


- do regressions, do regressions on subsets of data too 

- having descriptive statistics -- sometimes on 1 variable at a time, but usually between 2+ variables --> showing relationship on deeper level than a scatter plot (AKA regression, difference in means, etc.)

- 